\textbf{Microbiota predict \emph{Clostridium difficile} severity in germ-free mice colonized with human feces}

### 2017-01-06 submission/Manuscript.tex

Gnotobiotic or germ-free mouse models have been used for a range of
studies of CDI, including assessment of species-specific interactions
between \emph{C. difficile} and competing microbial community members
{[}11{]}, analysis of nutrient restriction {[}12{]}, \emph{in vivo}
transcriptomics of \emph{C. difficile} and examination of host immune
response to CDI {[}\textbf{???}{]}. Further, CDI therapeutics such as
antibiotics and fecal microbiota transplants have been tested
extensively in a gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model
of disease {[}13,14{]}. Pigs have a longer digestive tract with
components more similar to humans than mice and are typically infected
by strains typical in human infection {[}15{]}. However, the murine and
porcine microbiomes typically do not resemble those of the human gut.

The power of the gnotobiotic models to study CDI has been further
realized by first inoculating germ-free mice and piglets with human
stool microbes. In one study, germ-free piglets were acutely colonized
with human feces for one week and then treated with tigecycline. After
challenge with \emph{C. difficile} none of the antibiotic-treated
piglets succumbed to infection, while some of the untreated
human-colonized pigs did {[}13{]}. In another study, germ-free mice
colonized with human feces were bred over several generations to create
a cohort of mice with identical human-derived microbiomes {[}16{]}.
These mice were subsequently treated with a five-antibiotic cocktail to
induce dysbiosis and then were successfully colonized by \emph{C.
difficile} {[}16{]}. While informative, these studies were limited in
their use of only one human donor as input inoculum.

@@

\textbf{The microbiome predicts \emph{C. difficile} infection severity.}
As \emph{C. difficile} colonization levels can be predicted by the
microbiome composition on day 0, we postulated that severity of disease
could also be predicted from day 0. Using weight loss as a proxy for
severity, we built a Random Forest classification model based on
relative abundance to determine which community members were associated
with either mild or severe disease. We optimized the model using the
AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with
maximal Area Under the Curve (AUC) {[}17{]}, as used previously in our
group to classify colon cancer outcomes {[}18{]}. The optimized model
used X number of OTUs and identified several that predicted increased
weightloss and subsequent disease severity (Fig 4A, 4B). The populations
that were most commonly found in mice that succumbed to \emph{C.
difficile} infection were members of the \emph{Clostridiales},
\emph{Bacilliales}, \emph{Streptococcus} (Lactobacilles?) and
\emph{Bacteriodes}. (need more OTUs here, or number that were used for
model and then group by rank). The AUC was calculated using the
Out-Of-Bag (OOB) error for each sample (Fig 4A). Cross-validation using
10-fold, leave-one-mouse-out and leave-one-cage-out resulted in AUCs not
significantly different from the OOB (Fig 4A). This model reveals that
the levels of X OTUs on day 0 can predict disease outcome in mice with
human-associated microbiomes. (probably also going to want a
supplemental table for predictive OTUs for this and Fig 3)

@@

Here we colonized germ-free mice with a variety of human stool
communities and allowed these communities to colonize the mice for 2
weeks, prior to \emph{C. difficile} challenge. It is important to note
that the mouse gut environment is different from that of the human, and
as such exerts a selective pressure on the human stool microbial
community that stabilizes over the 2 weeks. Thus these mice cannot be
considered ``humanized'' but rather colonized by human-associated
microbes. Interestingly, clinical status was not associated with
proclivity to severe \emph{C. difficile} infection as some healthy donor
mice were susceptible. A surprising finding was that none of the
human-associated mouse communities were completely resistant to
colonization by \emph{C. difficile}. This is in contrast to numerous
reports of mouse microbiome studies {[}8{]} and the knowledge that in
general healthy humans without gut perturbations are resistant to
\emph{C. difficile} infection {[}Heinlen2010{]}. On the other hand, it
is known that humans, especially health care employees, can be
asymptomatic carriers of \emph{C. difficile} {[}19{]}. This may explain
the fact that the majority of our groups of mice were colonized by
\emph{C. difficile} but did not display outward signs of illness or
infection-associated weight loss (Fig 2).

@@

Further experiments will test if infecting susceptible communities with
weak strains allows the mice to survive. It has been recently reported
that carriage of a non-toxigenic strain of \emph{C. difficile} can
prevent additional colonization or recurrence by a pathogenic strain,
suggesting a role of niche-filling or immunity this pre-colonization
provides {[}20{]}. One interesting experiment would be to recapitulate
this finding in human-associated mouse microbiomes by making a
susceptible community resistant to infection. Further work with
human-associated communities in mice and mathematical modeling will
allow for uncovering of complex human-associated disease processes and
dynamics as they relate to human \emph{C. difficile} infection.

@@

\textbf{Human samples} Donor stool samples were chosen from a sample
bank collected from patients from October 2010 to November 2012 at the
University of Michigan. Sample collection methods and study was approved
by the Institutional Review Board at the University of Michigan. Stool
samples from University of Michigan Health System patients were
submitted to the microbiology lab for \emph{C. difficile} testing.
\emph{C. difficile} presence was measured using C.Diff Quik Chek
Complete Assay (Techlab). Samples from these patients were stored in
Cary-Blair medium at -80C. All patients consented to participation in
the study. Donor stool for this study was selected from one patient that
tested positive for \emph{C. difficile} infection and five patients that
presented with diarrhea and tested negative for \emph{C. difficile}.
Healthy donor stool was collected from adult volunteers living in the
Ann Arbor, MI area. 100ul of stool slurry was gavaged into each mouse.

\textbf{\emph{Clostridium difficile} and sample processing} For all
experiments, 100 spores of \emph{Clostridium difficile} were used to
orally infect mice as described previously {[}8{]}. \emph{C. difficile}
strain 431 is an 027 ribotype clinical isolate isolated from a patient
with severe CDI {[}21{]}. After gavage, the remaining inoculum was
diluted and plated to confirm infection dose. Bacteria were cultured
anaerobically on TCCFA plates and incubated at 37C {[}22{]}. During
infection, two fresh stool pellets were collected from each mouse daily.
One pellet was resuspended in anaerobic phosphate-buffered saline and
diluted and plated on TCCFA plates to enumerate \emph{C. difficile} CFU.
The other pellet was frozen at -80C until the end of the experiment.

\textbf{DNA isolation and sequencing} Bacterial DNA was isolated from
banked donor stool and each mouse stool pellet using the MO-BIO
PowerSoil DNA isolation kit. This DNA was used as template for
amplification of the V4 region of the 16S rRNA gene and fragments were
sequenced on an Illumina MiSeq as previously described {[}23{]}.

@@

13. Kim HB, Zhang Q, Sun X, Beamer G, Wang Y, Tzipori S. Beneficial
effect of oral tigecycline treatment on clostridium difficile infection
in gnotobiotic piglets. Antimicrobial Agents and Chemotherapy. American
Society for Microbiology; 2014;58: 7560--7564.
doi:\href{http://dx.doi.org/10.1128/aac.03447-14}{10.1128/aac.03447-14}

14. Diao H, Yan HL, Xiao Y, Yu B, Yu J, He J, et al. Intestinal
microbiota could transfer host gut characteristics from pigs to mice.
BMC Microbiology. Springer Nature; 2016;16.
doi:\href{http://dx.doi.org/10.1186/s12866-016-0851-z}{10.1186/s12866-016-0851-z}

15. Steele J, Feng H, Parry N, Tzipori S. Piglet models of acute or
chronic clostridium difficile illness. The Journal of Infectious
Diseases. Oxford University Press (OUP); 2010;201: 428--434.
doi:\href{http://dx.doi.org/10.1086/649799}{10.1086/649799}

16. Collins J, Auchtung JM, Schaefer L, Eaton KA, Britton RA. Humanized
microbiota mice as a model of recurrent clostridium difficile disease.
Microbiome. Springer Nature; 2015;3.
doi:\href{http://dx.doi.org/10.1186/s40168-015-0097-2}{10.1186/s40168-015-0097-2}

17. Calle ML, Urrea V, Boulesteix A-L, Malats N. AUC-RF: A new strategy
for genomic profiling with random forest. Human Heredity. S. Karger AG;
2011;72: 121--132.
doi:\href{http://dx.doi.org/10.1159/000330778}{10.1159/000330778}

18. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. Springer Nature; 2016;8.
doi:\href{http://dx.doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}

19. Furuya-Kanamori L, Marquess J, Yakob L, Riley TV, Paterson DL,
Foster NF, et al. Asymptomatic clostridium difficile colonization:
Epidemiology and clinical implications. BMC Infectious Diseases.
Springer Nature; 2015;15.
doi:\href{http://dx.doi.org/10.1186/s12879-015-1258-4}{10.1186/s12879-015-1258-4}

20. Gerding DN, Meyer T, Lee C, Cohen SH, Murthy UK, Poirier A, et al.
Administration of spores of nontoxigenic clostridium difficile strain m3
for prevention of recurrent c difficile infection. JAMA. American
Medical Association (AMA); 2015;313: 1719.
doi:\href{http://dx.doi.org/10.1001/jama.2015.3725}{10.1001/jama.2015.3725}

21. Carlson PE, Walk ST, Bourgis AE, Liu MW, Kopliku F, Lo E, et al. The
relationship between phenotype, ribotype, and clinical disease in human
clostridium difficile isolates. Anaerobe. Elsevier BV; 2013;24:
109--116.
doi:\href{http://dx.doi.org/10.1016/j.anaerobe.2013.04.003}{10.1016/j.anaerobe.2013.04.003}

22. Bliss DZ, Johnson S, Clabots CR, Savik K, Gerding DN. Comparison of
cycloserine-cefoxitin-fructose agar (CCFA) and taurocholate-CCFA for
recovery of clostridium difficile during surveillance of hospitalized
patients. Diagnostic Microbiology and Infectious Disease. Elsevier BV;
1997;29: 1--4.
doi:\href{http://dx.doi.org/10.1016/s0732-8893(97)00113-2}{10.1016/s0732-8893(97)00113-2}

23. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq illumina sequencing
platform. Applied and Environmental Microbiology. American Society for
Microbiology; 2013;79: 5112--5120.
doi:\href{http://dx.doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}

>>>>

Gnotobiotic or germ-free mouse models have been used for a range of
studies of CDI, including assessment of species-specific interactions
between \emph{C. difficile} and competing microbial community members
{[}11{]}, analysis of nutrient restriction {[}12{]}, \emph{in vivo}
transcriptomics of \emph{C. difficile} and examination of host immune
response to CDI {[}13{]}. Further, CDI therapeutics such as antibiotics
and fecal microbiota transplants have been tested extensively in a
gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease
{[}14,15{]}. Pigs have a longer digestive tract with components more
similar to humans than mice and are typically infected by strains
typical in human infection {[}16{]}. However, the murine and porcine
microbiomes typically do not resemble those of the human gut.

The power of the gnotobiotic models to study CDI has been further
realized by first inoculating germ-free mice and piglets with human
stool microbes. In one study, germ-free piglets were acutely colonized
with human feces for one week and then treated with tigecycline. After
challenge with \emph{C. difficile} none of the antibiotic-treated
piglets succumbed to infection, while some of the untreated
human-colonized pigs did {[}14{]}. In another study, germ-free mice
colonized with human feces were bred over several generations to create
a cohort of mice with identical human-derived microbiomes {[}17{]}.
These mice were subsequently treated with a five-antibiotic cocktail to
induce dysbiosis and then were successfully colonized by \emph{C.
difficile} {[}17{]}. While informative, these studies were limited in
their use of only one human donor as input inoculum.

@@

\textbf{The microbiome predicts \emph{C. difficile} infection severity.}
As \emph{C. difficile} colonization levels can be predicted by the
microbiome composition on day 0, we postulated that severity of disease
could also be predicted from day 0. Using weight loss as a proxy for
severity, we built a Random Forest classification model based on
relative abundance to determine which community members were associated
with either mild or severe disease. We optimized the model using the
AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with
maximal Area Under the Curve (AUC) {[}18{]}, as used previously in our
group to classify colon cancer outcomes {[}19{]}. The optimized model
used X number of OTUs and identified several that predicted increased
weightloss and subsequent disease severity (Fig 4A, 4B). The populations
that were most commonly found in mice that succumbed to \emph{C.
difficile} infection were members of the \emph{Clostridiales},
\emph{Bacilliales}, \emph{Streptococcus} (Lactobacilles?) and
\emph{Bacteriodes}. (need more OTUs here, or number that were used for
model and then group by rank). The AUC was calculated using the
Out-Of-Bag (OOB) error for each sample (Fig 4A). Cross-validation using
10-fold, leave-one-mouse-out and leave-one-cage-out resulted in AUCs not
significantly different from the OOB (Fig 4A). This model reveals that
the levels of X OTUs on day 0 can predict disease outcome in mice with
human-associated microbiomes. (probably also going to want a
supplemental table for predictive OTUs for this and Fig 3)

@@

Here we colonized germ-free mice with a variety of human stool
communities and allowed these communities to colonize the mice for 2
weeks, prior to \emph{C. difficile} challenge. It is important to note
that the mouse gut environment is different from that of the human, and
as such exerts a selective pressure on the human stool microbial
community that stabilizes over the 2 weeks. Thus these mice cannot be
considered ``humanized'' but rather colonized by human-associated
microbes. Interestingly, clinical status was not associated with
proclivity to severe \emph{C. difficile} infection as some healthy donor
mice were susceptible. A surprising finding was that none of the
human-associated mouse communities were completely resistant to
colonization by \emph{C. difficile}. This is in contrast to numerous
reports of mouse microbiome studies {[}8{]} and the knowledge that in
general healthy humans without gut perturbations are resistant to
\emph{C. difficile} infection {[}20{]}. On the other hand, it is known
that humans, especially health care employees, can be asymptomatic
carriers of \emph{C. difficile} {[}21{]}. This may explain the fact that
the majority of our groups of mice were colonized by \emph{C. difficile}
but did not display outward signs of illness or infection-associated
weight loss (Fig 2).

@@

It has been recently reported that carriage of a non-toxigenic strain of
\emph{C. difficile} can prevent additional colonization or recurrence by
a pathogenic strain, suggesting a role of niche-filling or immunity this
pre-colonization provides {[}22{]}. One interesting experiment would be
to recapitulate this finding in human-associated mouse microbiomes by
making a susceptible community resistant to infection. Further work with
human-associated communities in mice and mathematical modeling will
allow for uncovering of complex human-associated disease processes and
dynamics as they relate to human \emph{C. difficile} infection.

@@

\textbf{Human samples} Donor stool samples were chosen from a sample
bank collected from patients from October 2010 to November 2012 at the
University of Michigan Health System. Sample collection methods and the
study were approved by the Institutional Review Board at the University
of Michigan. Stool samples from University of Michigan Health System
patients were submitted to the microbiology lab for \emph{C. difficile}
testing. \emph{C. difficile} presence was measured using C.Diff Quik
Chek Complete Assay (Techlab). Fecal samples from these patients were
stored in Cary-Blair medium at -80C. All patients consented to
participation in the study. Donor stool was selected from one patient
that tested positive for \emph{C. difficile} infection and five patients
that presented with diarrhea and tested negative for \emph{C.
difficile}. Healthy donor stool was collected from adult volunteers
living in the Ann Arbor, MI area. 100ul of stool slurry was gavaged into
each mouse.

\textbf{\emph{Clostridium difficile} and sample processing} For all
experiments, 100 spores of \emph{Clostridium difficile} were used to
orally infect mice as described previously {[}8{]}. \emph{C. difficile}
strain 431 is an 027 ribotype clinical isolate isolated from a patient
with severe CDI {[}23{]}. After gavage, the remaining inoculum was
diluted and plated to confirm infection dose. Bacteria were cultured
anaerobically on TCCFA plates and incubated at 37C {[}24{]}. During
infection, two fresh stool pellets were collected from each mouse daily.
One pellet was resuspended in anaerobic phosphate-buffered saline and
diluted and plated on TCCFA plates to enumerate \emph{C. difficile} CFU.
The other pellet was frozen at -80C until the end of the experiment.

\textbf{DNA isolation and sequencing} Bacterial DNA was isolated from
banked donor stool and each mouse stool pellet using the MO-BIO
PowerSoil DNA isolation kit. This DNA was used as template for
amplification of the V4 region of the 16S rRNA gene and fragments were
sequenced on an Illumina MiSeq as previously described {[}25{]}.

@@

13. Pawlowski SW, Calabrese G, Kolling GL, Freire R, AlcantaraWarren C,
Liu B, et al. Murine model of clostridium difficile infection with aged
gnotobiotic c57BL/6 mice and a BI/NAP1 strain. The Journal of Infectious
Diseases. Oxford University Press (OUP); 2010;202: 1708--1712.
doi:\href{http://dx.doi.org/10.1086/657086}{10.1086/657086}

14. Kim HB, Zhang Q, Sun X, Beamer G, Wang Y, Tzipori S. Beneficial
effect of oral tigecycline treatment on clostridium difficile infection
in gnotobiotic piglets. Antimicrobial Agents and Chemotherapy. American
Society for Microbiology; 2014;58: 7560--7564.
doi:\href{http://dx.doi.org/10.1128/aac.03447-14}{10.1128/aac.03447-14}

15. Diao H, Yan HL, Xiao Y, Yu B, Yu J, He J, et al. Intestinal
microbiota could transfer host gut characteristics from pigs to mice.
BMC Microbiology. Springer Nature; 2016;16.
doi:\href{http://dx.doi.org/10.1186/s12866-016-0851-z}{10.1186/s12866-016-0851-z}

16. Steele J, Feng H, Parry N, Tzipori S. Piglet models of acute or
chronic clostridium difficile illness. The Journal of Infectious
Diseases. Oxford University Press (OUP); 2010;201: 428--434.
doi:\href{http://dx.doi.org/10.1086/649799}{10.1086/649799}

17. Collins J, Auchtung JM, Schaefer L, Eaton KA, Britton RA. Humanized
microbiota mice as a model of recurrent clostridium difficile disease.
Microbiome. Springer Nature; 2015;3.
doi:\href{http://dx.doi.org/10.1186/s40168-015-0097-2}{10.1186/s40168-015-0097-2}

18. Calle ML, Urrea V, Boulesteix A-L, Malats N. AUC-RF: A new strategy
for genomic profiling with random forest. Human Heredity. S. Karger AG;
2011;72: 121--132.
doi:\href{http://dx.doi.org/10.1159/000330778}{10.1159/000330778}

19. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. Springer Nature; 2016;8.
doi:\href{http://dx.doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}

20. Heinlen L, Ballard JD. Clostridium difficile infection. The American
Journal of the Medical Sciences. Elsevier BV; 2010;340: 247--252.
doi:\href{http://dx.doi.org/10.1097/maj.0b013e3181e939d8}{10.1097/maj.0b013e3181e939d8}

21. Furuya-Kanamori L, Marquess J, Yakob L, Riley TV, Paterson DL,
Foster NF, et al. Asymptomatic clostridium difficile colonization:
Epidemiology and clinical implications. BMC Infectious Diseases.
Springer Nature; 2015;15.
doi:\href{http://dx.doi.org/10.1186/s12879-015-1258-4}{10.1186/s12879-015-1258-4}

22. Gerding DN, Meyer T, Lee C, Cohen SH, Murthy UK, Poirier A, et al.
Administration of spores of nontoxigenic clostridium difficile strain m3
for prevention of recurrent c difficile infection. JAMA. American
Medical Association (AMA); 2015;313: 1719.
doi:\href{http://dx.doi.org/10.1001/jama.2015.3725}{10.1001/jama.2015.3725}

23. Carlson PE, Walk ST, Bourgis AE, Liu MW, Kopliku F, Lo E, et al. The
relationship between phenotype, ribotype, and clinical disease in human
clostridium difficile isolates. Anaerobe. Elsevier BV; 2013;24:
109--116.
doi:\href{http://dx.doi.org/10.1016/j.anaerobe.2013.04.003}{10.1016/j.anaerobe.2013.04.003}

24. Bliss DZ, Johnson S, Clabots CR, Savik K, Gerding DN. Comparison of
cycloserine-cefoxitin-fructose agar (CCFA) and taurocholate-CCFA for
recovery of clostridium difficile during surveillance of hospitalized
patients. Diagnostic Microbiology and Infectious Disease. Elsevier BV;
1997;29: 1--4.
doi:\href{http://dx.doi.org/10.1016/s0732-8893(97)00113-2}{10.1016/s0732-8893(97)00113-2}

25. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq illumina sequencing
platform. Applied and Environmental Microbiology. American Society for
Microbiology; 2013;79: 5112--5120.
doi:\href{http://dx.doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}
